---
title: "1b_reading_arrest_data"
author: "Brit Henderson"
date: "Last Updated: `r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

# Objective
```{r directories}
rawdatdir <- "/data/share/xproject/Training/Practice/henderson/Dissertation/raw"
rdadir <- "/data/share/xproject/Training/Practice/henderson/Dissertation/rda/Intermediate"

```

```{r functions}

```

```{r packages}
# # Import packages
library(Hmisc)
library(tidyverse) # suite of packages in for manipulating/tidying/visualizing, etc. data
library(lubridate) # manipulating dates
library(readxl)  # read in excel files
library(knitr) # for RMD formatting
library(kableExtra) # to output tables as images
library(fastDummies) # easy creation of dummy variables 



```

# Create variables for index charges
```{r index}
#Read in index charge data and create variables
dat.index1 <- readRDS(here::here("raw/Main Arrests Data/ucr_arrests_yearly_index_crimes_age_1974_2019.rds"))

dat.index2 <- dat.index1 %>% 
  select(-contains("female"), 
         -contains("male"), 
         -contains("juv"),
         -contains("_15"),
         -contains("_13_14"),
         -contains("_under_10"),
         -contains("10_12"),
         -contains("_16"),
         -contains("_17"))  
  
  
# Create variable capturing young adult total index charges
    dat.index3 <- dat.index2 %>% 
      mutate(young_adult_index_tot = select(dat.index2, 
                                        contains("_18"),
                                  contains("_19"),
                                  contains("_20"),
                                  contains("_21"),
                                  contains("_22"),
                                  contains("_23"),
                                  contains("_24")) %>% 
      rowSums(na.rm = TRUE))

# checking young adult variable
dat.check1 = filter(dat.index3, year==2018, fips_state_county_code == "02020", agency_name == "anchorage") %>% 
  select( year, fips_state_county_code, young_adult_index_tot,
                                        contains("_18"),
                                  contains("_19"),
                                  contains("_20"),
                                  contains("_21"),
                                  contains("_22"),
                                  contains("_23"),
                                  contains("_24"))



# Create variable capturing other adult total index charges
    dat.index4 <- dat.index3 %>% 
      mutate(other_adult_index_tot = select(dat.index3, 
                                      contains("_25"),
                                  contains("_30"),
                                  contains("_35"),
                                  contains("_40"),
                                  contains("_45"),
                                  contains("_50"),
                                  contains("_55"),
                                  contains("_60"),
                                  contains("_over_64")) %>% 
      rowSums(na.rm = TRUE))

# checking other adult variable
dat.check2 = filter(dat.index4, year==2018, fips_state_county_code == "02020", agency_name == "anchorage") %>% 
  select( year, fips_state_county_code, other_adult_index_tot,
                                        contains("_25"),
                                  contains("_30"),
                                  contains("_35"),
                                  contains("_40"),
                                  contains("_45"),
                                  contains("_50"),
                                  contains("_55"),
                                  contains("_60"),
                                  contains("_over_64"))

# create measure for all adults and dropped extra columns

dat.index5 <- dat.index4 %>% 
  select(-contains("_18"),
         -contains("_19"),
         -contains("_20"),
         -contains("_21"),
         -contains("_22"),
         -contains("_23"),
         -contains("_24"),
         -contains("_25"),
         -contains("_30"),
         -contains("_35"),
         -contains("_40"),
         -contains("_45"),
         -contains("_50"),
         -contains("_55"),
         -contains("_60"),
         -contains("_over_64"),
         -contains("_tot_"),
         -contains("num_")) %>% 
  mutate(all_adult_index_tot = young_adult_index_tot + other_adult_index_tot)
  
saveRDS(dat.index5, file = file.path(rdadir, "int_index.rds") )

```


# Create variables for property charges
```{r property}
#Read in property charge data and create variables
dat.prop1 <- readRDS(here::here("raw/Main Arrests Data/ucr_arrests_yearly_alcohol_or_property_crimes_age_1974_2019.rds"))

dat.prop2 <- dat.prop1 %>% 
  select(-contains("female"), 
         -contains("male"), 
         -contains("juv"),
         -contains("_15"),
         -contains("_13_14"),
         -contains("_under_10"),
         -contains("10_12"),
         -contains("_16"),
         -contains("_17"))  
  
  
# Create variable capturing young adult total prop charges
    dat.prop3 <- dat.prop2 %>% 
      mutate(young_adult_prop_tot = select(dat.prop2, 
                                        contains("_18"),
                                  contains("_19"),
                                  contains("_20"),
                                  contains("_21"),
                                  contains("_22"),
                                  contains("_23"),
                                  contains("_24")) %>% 
      rowSums(na.rm = TRUE))

# checking young adult variable
dat.check1 = filter(dat.prop3, year==2018, fips_state_county_code == "02020", agency_name == "anchorage") %>% 
  select( year, fips_state_county_code, young_adult_prop_tot,
                                        contains("_18"),
                                  contains("_19"),
                                  contains("_20"),
                                  contains("_21"),
                                  contains("_22"),
                                  contains("_23"),
                                  contains("_24"))



# Create variable capturing other adult total prop charges
    dat.prop4 <- dat.prop3 %>% 
      mutate(other_adult_prop_tot = select(dat.prop3, 
                                      contains("_25"),
                                  contains("_30"),
                                  contains("_35"),
                                  contains("_40"),
                                  contains("_45"),
                                  contains("_50"),
                                  contains("_55"),
                                  contains("_60"),
                                  contains("_over_64")) %>% 
      rowSums(na.rm = TRUE))

# checking other adult variable
dat.check2 = filter(dat.prop4, year==2018, fips_state_county_code == "02020", agency_name == "anchorage") %>% 
  select( year, fips_state_county_code, other_adult_prop_tot,
                                        contains("_25"),
                                  contains("_30"),
                                  contains("_35"),
                                  contains("_40"),
                                  contains("_45"),
                                  contains("_50"),
                                  contains("_55"),
                                  contains("_60"),
                                  contains("_over_64"))

# create measure for all adults and dropped extra columns

dat.prop5 <- dat.prop4 %>% 
  select(-contains("_18"),
         -contains("_19"),
         -contains("_20"),
         -contains("_21"),
         -contains("_22"),
         -contains("_23"),
         -contains("_24"),
         -contains("_25"),
         -contains("_30"),
         -contains("_35"),
         -contains("_40"),
         -contains("_45"),
         -contains("_50"),
         -contains("_55"),
         -contains("_60"),
         -contains("_over_64"),
         -contains("_tot_"),
         -contains("num_")) %>% 
  mutate(all_adult_prop_tot = young_adult_prop_tot + other_adult_prop_tot)
  
saveRDS(dat.prop5, file = file.path(rdadir, "int_prop.rds") )

```


#Merge with county fips crosswalk file
```{r}
#Read in county fips crosswalk file
fips_xwalk <- read.csv(here::here("raw/Census Data/FIPS_code_xwalk.csv"), colClasses = c(FIPS = "character"))

# Add leading zeros to Fips codes in xwalk
fips_xwalk$fips_clean <- str_pad(fips_xwalk$FIPS, 5, pad = "0")
  #paste0("0",fips_xwalk$FIPS) - this code could be useful if everything in column needs leading zero. otherwise requires conditioning
  

#In population data set, create composite fips code variable that includes county fips code and state fips code
dat5$fips_clean <- paste(dat5$state.fips,dat5$county.fips,sep="")

dat6 <- merge(dat5,fips_xwalk,by="fips_clean", all = TRUE)
count(filter(dat6,is.na(State)==1 & is.na(state)==0))
count(filter(dat6,is.na(State)==0 & is.na(state)==1))
#count(filter(dat6,is.na(State)==0 & is.na(state)==1, !(State %in% c('PR','AS','VI','MP','GU'))))

as.data.frame(filter(dat6,is.na(State)==0 & is.na(state)==1) %>% select(fips_clean,FIPS, Name, State))  
```


#summary statistics for both data sets
```{r}
summary(dat6)
describe(dat6)
#There are 16 county-years observations that did not have any young adults
# The largest number of young adults is 1,097,486

dat6 %>% filter(is.na(State) ==1) %>%  #select(unique())
```
There are `r sum(is.na(dat5$n_young_adults)==1)` county-year observations that did not have any young adults.
There are `r count(filter(dat6,is.na(State)==1 & is.na(state)==0))` observations that are in the FIPS crosswalk but not in the population data set.
There are `r count(filter(dat6,is.na(State)==0 & is.na(state)==1))` observations that are in the population data set but not in the FIPS crosswalk. Most of these were for geographies in Puerto Rico, American Samoa, U.S. Virgin Islands, Northern Mariana Islands, and Guam. There were also
`r count(filter(dat6,is.na(State)==0 & is.na(state)==1, State = "VA")` records for Virginia and 1 record for Yellowstone National Park in Montana that did not match to any FIPS codes. This is fine.  
The largest number of young adults is 1,097,486.

#Cleaning final data set
```{r}
dat7 <- dat6 %>% select(-state,-FIPS)
```


#Save files to type rda

```{r}
save(dat4, file = file.path(rdadir, ""))
save(dat7, file = file.path(rdadir, ""))
```

